---
title: "VEIN model"
author: "Sergio Ibarra-Espinosa"
date: "22 de abril de 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Vein Package

**V**ehicular **E**missions **IN**ventory **(VEIN)** model.  

Read the DESCRIPTION file. You need several packages to run **vein**.

- sp
- raster
- rgdal
- rgeos

Also, if you want to properly run the demo, install these packages:

- ggplot2
- DiagrammeR
- RColorBrewer


```{r eval=F}
library(devtools)
install_github("ibarraespinosa/vein")
library(vein)
demo("VEIN")
```

I'm currently working in some classes.

I've created the following classes:

- Vehicles
- Speed
- EmissionFactors
- EmissionFactorsList
- Emissions
- EmissionsArray
- EmissionsList

Each has a constructor function with the same name that return the class. I'm
trying to create methods for the generic functions 'print', 'summary' and 
'plot'.

The checking --as-cran `devtools::check(args = c('--as-cran'))` of the package shows 1 warning and 2 notes about the 
**"S3 methods shown with full name in documentation object"**. I'm fixing that.
Anyway, the demo runs fine but Classes are not methods of classes not fully
implemented.

Isseus here please https://github.com/ibarraespinosa/vein/issues/

I hope you enjoy it
