---
title: "Understanding speed-dependent inventories with emis_speed"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding speed-dependent inventories with emis_speed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

This article explains, step by step, how VEIN builds an hourly, per-street
inventory with **speed-dependent** emission factors, and how `emis_speed()`
makes it fast enough for country-scale networks. Every chunk runs; you can copy
and paste them.

## 1. The idea

A local emission factor (for example the CETESB factors used in Brazil) is a
single number per vehicle age, measured over a driving cycle. Real emissions
depend on the average speed of the link. The EMEP/EEA guidebook provides
*speed curves* for the same technologies. VEIN combines both: the EMEP/EEA
curve gives the **shape** as a function of speed, rescaled so that at the speed
of the local driving cycle it reproduces the **local** emission factor.

Doing this for every street and every hour used to be slow because the curve was
interpreted in R once per street, hour and age. `emis_speed()` compiles the
curves and evaluates them in a C/OpenMP kernel.

## 2. Local emission factors by age

`ef_cetesb()` returns one emission factor per age. `full = TRUE` also gives the
Euro equivalence standard of each age, which we need to pick the right curve.

```{r}
library(vein)

CO <- ef_cetesb(p = "CO", veh = "PC_G", year = 2018, agemax = 40, full = TRUE)
head(CO[, c("Age", "Year", "Proconve_PC", "EqEuro_PC", "CO")], 8)
```

More than one call to `ef_cetesb()` with the same `year`/`scale` is cheap: the
large, pollutant-independent CETESB table is cached internally.

## 3. Scaling a curve to the driving cycle (SDC)

`ef_ldv_scaled()` (light vehicles) and `ef_hdv_scaled()` (heavy vehicles
and buses) return one function per age. Each function is
`EF_a(V) = k_a * curve_a(V)` where `k_a = EF_local[a] / curve_a(SDC)`, so that
at the driving-cycle speed `SDC` we recover the local factor exactly.

```{r}
lef <- ef_ldv_scaled(
  dfcol = CO$CO, SDC = 34.12,
  v = "PC", t = "4S", cc = "<=1400", f = "G", eu = CO$EqEuro_PC, p = "CO"
)

length(lef) # one function per age
# at the driving-cycle speed the scaled function equals the local factor
c(local = as.numeric(CO$CO[1]), scaled = as.numeric(lef[[1]](34.12)))
```

Once scaled, the factor depends on speed. Ages with a newer Euro standard have a
different curve, so the shape also changes with age:

```{r}
V <- seq(0, 130, by = 10)
round(sapply(c(1, 10, 20, 40), function(a) as.numeric(lef[[a]](V))), 3)
```

## 4. What `emis_speed()` consumes: compiled programs

Evaluating R closures for millions of street-hours is slow. Instead we compile
each curve to a tiny stack machine and **group the ages that share the same
equation**. `programs_only = TRUE` returns just that bundle.

```{r}
prog <- ef_ldv_scaled(
  dfcol = CO$CO, SDC = 34.12,
  v = "PC", t = "4S", cc = "<=1400", f = "G", eu = CO$EqEuro_PC, p = "CO",
  programs_only = TRUE
)
class(prog)
c(ages = prog$n, groups = prog$G) # 40 ages collapse into a few equations
str(prog, max.level = 1)
```

The important fields:

- `code`, `consts`: the RPN bytecode of each distinct equation.
- `cofs`, `minv`, `maxv`: coefficients `a..f` and the speed clamp per group.
- `gid`: for each age, which group (equation) it uses.
- `kk_age`: the per-age scaling constant `k_a`.

## 5. Speeds per street and hour

Emissions need a speed for each link and each hour of the week. `temp_fact()`
expands the traffic flow with a temporal profile and `netspeed()` turns flow
into speed with a BPR function.

```{r}
data(net)
data(pc_profile)

total <- net$ldv + net$hdv
flow  <- temp_fact(total, pc_profile)                                  # 168 h
speed <- netspeed(flow, net$ps, net$ffs, net$capacity, net$lkm, alpha = 1)
dim(speed) # streets x (24 h * 7 days)
```

## 6. Running the inventory

`veh` is a matrix of vehicle flow (veh/h) by street (rows) and age (columns).
`emis_speed()` returns the emissions by **street x hour** and, with
`by_age = TRUE`, the totals by **age x hour**.

```{r}
A <- 40
veh <- matrix(as.numeric(net$ldv) / A, nrow = nrow(net), ncol = A)

E <- emis_speed(
  veh = veh, lkm = net$lkm, ef = lef,
  speed = speed, profile = pc_profile,
  by_age = TRUE, nt = 2
)

dim(E$streets) # streets x hours
dim(E$veh)     # ages x hours
sum(E$streets) # grams per hour over the network
```

The units of `E$streets` are g/h because `veh` is veh/h, `lkm` is km and the
emission factor is g/km.

## 7. It matches the reference implementation

`emis_speed()` is a drop-in replacement for the speed branch of `emis()`; the
results agree to machine precision.

```{r}
Eref <- emis(
  veh = veh, lkm = net$lkm, ef = lef,
  speed = speed, profile = pc_profile,
  simplify = TRUE, agemax = A
)

max(abs(apply(Eref, c(1, 3), sum) - E$streets))
```

## 8. Under the hood

1. **Compile.** Each `Y` equation in `sysdata$ldv` / `sysdata$hdv_*` is parsed
   once into RPN bytecode (`compile_ef_expr()`); the result is cached.
2. **Group.** Ages with identical equation and coefficients (usually several
   years within one Euro standard) are merged, so a curve is evaluated once per
   street-hour instead of once per age. For this example the 40 ages need only
   `r prog$G` equations.
3. **Evaluate.** `src/e_speed.c` runs the bytecode in a tight OpenMP loop and
   accumulates the totals directly, never materialising the full
   street x age x hour array.

A quick timing comparison:

```{r}
S <- 1000
vh <- matrix(50, S, A)
sp <- Speed(matrix(40, S, H <- 168))
pf <- matrix(1, 24, 7)
lkm1 <- units::set_units(rep(1, S), "km")

t_ref <- system.time(
  E1 <- emis(veh = vh, lkm = lkm1, ef = lef, speed = sp, profile = pf,
             simplify = TRUE, agemax = A)
)["elapsed"]
t_new <- system.time(
  E2 <- emis_speed(veh = vh, lkm = lkm1, ef = lef, speed = sp, profile = pf,
                   agemax = A, nt = 2)
)["elapsed"]
round(c(emis = t_ref, emis_speed = t_new), 3)
```

## 9. Country-scale workflow: the `brazil_bu_speed` project

The complete workflow is packaged as a project you can download and run:

```{r, eval = FALSE}
get_project(directory = tempdir(), case = "brazil_bu_speed")
```

Key files:

- `config/ef_speed_mapping.R` maps every CETESB category (e.g. `TRUCKS_M_D`)
  to an EMEP/EEA category (`v`, `t`, `cc`/`g`, `f`) and the Euro column of
  `ef_cetesb()`. It also defines `ef_cetesb_speed()`, the scaled-EF builder.
- `scripts/speed.R` computes the 168 hourly speeds with `netspeed()`.
- `scripts/exhaust_speed.R` loops over categories and pollutants, runs
  `emis_speed()` and falls back to a constant emission factor when no EMEP
  curve applies.
- `scripts/post_speed.R` grids the hourly street emissions and writes totals.

## 10. Tips and pitfalls

- **Not every curve is usable.** For some pollutant/Euro combinations the EMEP
  curve is zero at `SDC` (for example pre-Euro PM), so the scaling constant is
  not finite. `emis_speed()` stops with a clear message; the project detects
  this and uses the constant CETESB factor instead.
- **Threads.** `nt` controls OpenMP threads. The default is half of
  `check_nt()`, which works well when R and the kernel must share the machine.
- **Memory.** For huge networks the per-street result is only
  `streets x hours` (e.g. 200 000 x 168 is about 270 MB), because ages are
  summed in C. Ask for `by_age = TRUE` to also get the small `ages x hours`
  totals.
- **Speed layout.** The columns of `speed` must match the flattened
  `profile` (hours of day 1, then day 2, ...). `temp_fact()` produces this order.
