---
title: "Speed-dependent emissions with emis_speed"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Speed-dependent emissions with emis_speed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Why speed-dependent emission factors?

Local emission factors (for instance the CETESB values used in Brazil) are
usually measured over a driving cycle and reported as a single number per age.
The EMEP/EEA guidebook provides **speed curves** for the same technologies.
`ef_ldv_scaled()` and `ef_hdv_scaled()` combine both: the EMEP/EEA curve gives
the shape as a function of the average speed, and it is rescaled so that at the
speed of the local driving cycle (SDC) it reproduces the local emission factor.

Estimating an inventory for many streets and hours with these curves used to be
very slow, because the R equations were interpreted once per street, hour and
age. `emis_speed()` compiles the equations to a tiny stack machine and
evaluates them in an OpenMP C kernel (`src/e_speed.c`), grouping the ages that
share the same equation.

## Scaling a local emission factor by speed

```{r}
library(vein)
library(units)

# local (CETESB) emission factors for passenger cars, 40 ages
CO <- ef_cetesb(p = "CO", veh = "PC_G", year = 2018, agemax = 40, full = TRUE)

# scale the EMEP/EEA curve so that at SDC = 34.12 km/h it gives the local EF
lef <- ef_ldv_scaled(
  dfcol = CO$CO,
  v = "PC", t = "4S", cc = "<=1400", f = "G",
  eu = CO$EqEuro_PC, p = "CO"
)
length(lef)
```

## Running the inventory

`emis_speed()` needs the vehicle flow by age (`veh`), the link lengths (`lkm`),
the number of hours and speeds. It returns the emissions by street and hour and,
optionally, the totals by age and hour.

```{r}
S <- 200 # streets
A <- 40  # ages
H <- 24  # hours

veh <- matrix(100, nrow = S, ncol = A)
lkm <- set_units(rep(1, S), "km")
speed <- Speed(matrix(rep(c(10, 30, 50, 80), length.out = S * H), S, H))
profile <- matrix(1, 24, 1)

E <- emis_speed(
  veh = veh,
  lkm = lkm,
  ef = lef,
  speed = speed,
  profile = profile,
  by_age = TRUE,
  nt = 2
)

# emissions by street and hour
dim(E$streets)
# total emissions by age and hour
dim(E$veh)
round(sum(E$streets), 1)
```

The result is equivalent to the reference implementation `emis()` (to machine
precision) but considerably faster, which makes country-scale hourly
inventories feasible. The detailed workflow, with CETESB emission factors and a
168 h week, is implemented in the `brazil_bu_speed` project:

```{r, eval = FALSE}
get_project(directory = tempdir(), case = "brazil_bu_speed")
```
