---
title: "Basics for running VEIN"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basics for running VEIN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This vignette shows how to install and run VEIN for people who **do not know R**
  
  ## Install VEIN
  
  You can install VEIN with 

- GitLab: `remotes::install_gitlab("ibarraespinosa/vein")`
- GitHub: `remotes::install_github("atmoschem/vein")`
- or CRAN: `install.packages("vein")`

GitHub and GitLab are more updated

Then, use vein

```{r}
library(vein)
```


## Download a project

Check the documentation of get_project

```{r}
?get_project

```

or [here](https://atmoschem.github.io/vein/reference/get_project.html)

Choose a name, for instance, "awesome_city" 

```{r, eval = F}
get_project(directory = "awesome_city")
system("tree  awesome_city")

```
The structure is:
  
```
awesome_city
├── config
│   ├── clean.R
│   ├── config.R
│   ├── inventory.xlsx
│   └── packages.R
├── main.R
├── main.Rproj
├── network
│   ├── net.gpkg
│   └── net.rds
├── scripts
│   ├── evaporatives.R
│   ├── exhaust.R
│   ├── fuel_eval.R
│   ├── net.R
│   ├── pavedroads.R
│   ├── plots.R
│   ├── post.R
│   ├── traffic.R
│   └── wrf.R
└── wrf
└── wrfinput_d02

```

You have to open the file `main.Rproj` with Rstudio and then open and run `main.R`

To run `main.R` you will need these extra packages:

- ggplot2
- readxl
- eixport (If you plan to generate WRF Chem emissions file)

If you do not have them already, you can install:


```{r, eval = F}
install.packages(c("ggplot2", "readxl", "eixport"))
```

## Too complicated? Watch a YouTube video

in English:

<iframe width="560" height="315" src="https://www.youtube.com/embed/tHSWIjg26vg" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

in Portuguese:

<iframe width="560" height="315" src="https://www.youtube.com/embed/6-07Y0Eimng" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

[here](https://youtu.be/6-07Y0Eimng)

